Substructure Boosts to Dark Matter Annihilation from Sommerfeld Enhancement

### 2009-04-18 tex/response.txt

- I slightly rewrote the paragraph containing Eq. 20 to include the
definition of the virial radius;

- I fixed a type which the referee found in the paragraph below Eq. 39;

- I changed the color of the label in figures 1, 2, 5, 6, and 7 from
  black to white;

- I enlarged figures 9, 10, 12, 14, arranging the two panels now below
  each other. I updated the text and the caption of figure 10 to
  reflect this change.

- In the conclusion I elaborated a fit further on the implications of
  the possibility that the Galactic halo consists for a large part of
  cold dark matter streams, adding a reference to a paper by Afshordi,
  Mohayaee, and Bertschinger which discusses boosts due to such tidal
  debris.

- I also updated all of the references to papers which have gotton
  published since I first submitted this paper.

In addition to these I made a number of cosmetic changes and slight
rewordings to the text to improve the clarity of the exposition, the
most important of which are:

- Some rewordings in the abstract (including spelling out the proper
  name of the Fermi telescope, and emphasizing more clearly what is
  new in the present paper);

- I renamed Sec. III A "Some details of the particle physics model"
  instead of "Some details of the particle physics"

- I moved up the sentence in the paragraph above Eq. 13 which states
  that the DM annihilation flux depends on the density squared to the
  beginning of that section, such that the point of discussing the
  dark matter profile at length is clear at the start of the section
  on the shape of the profile.

### 2009-04-18 tex/somm.tex
sommboost: new version

\begin{abstract}
I discuss the implications of the recently introduced Sommerfeld
enhancement of the dark matter annihilation cross section for the
detection of subhalos in the Galactic halo.  In addition to the boost
to the dark matter annihilation cross section from the high densities
of these subhalos with respect to the main halo, an additional boost
due to Sommerfeld enhancement results from the fact that they are
kinematically colder than the Galactic halo. If we further believe the
generic prediction of \CDM\ that in each subhalo there is an abundance
of substructure which is approximately self-similar to that of the
Galactic halo, then additional boosts coming from the density
enhancements of these small substructures and their small velocity
dispersions enhance the dark matter annihilation cross section even
further. I show that very large boost factors ($10^5$ -- $10^9$) are
obtained in a large class of models. The implications of these boost
factors for the detection of dark matter annihilation from dwarf
Spheroidal galaxies in the Galactic halo are such that they outshine
the background gamma-ray flux and should be detectable
by \Fermi.
\end{abstract}

@@

One exciting possibility is that dark matter self-annihilates at a
level observable in the Galactic halo today. Detecting this
self-annihilation seems unlikely considering the standard
self-annihilation cross section --- $\sigmaannv_{\mathrm{ann}}
\lesssim 3 \times 10^{-26} $ cm$^3$ s$^{-1}$ for consistency with the
observed relic density. However, there are various mechanisms to boost
the annihilation signal to a value larger by a few orders of
magnitude, e.g., clumpiness of the dark halo
\cite{1993ApJ...411..439S,1999PhRvD..59d3506B,2008A&A...479..427L,2008ApJ...686..262K}
or a combination of coannihilations and Sommerfeld enhancement
\cite{2005PhRvD..72j3521P,2008arXiv0812.0360L}. The possibility of
detecting the \DM\ annihilation signal has attracted much attention
recently because of the anomalous excesses of high energy electrons
and positrons reported by two different experimental groups: the
Payload for Antimatter Exploration and Light-nuclei Astrophysics
(PAMELA) satellite, which saw a sharp upturn in the positron fraction
$e^+/(e^++e^-)$ from 10 -- 100 GeV \cite{Adriani:2008zr}, and the
Advanced Thin Ionization Calorimeter (ATIC) group, which reported an
excess in the combined number of electrons and positrons at energies
500 -- 800 GeV \cite{2008Natur.456..362C}. The PAMELA result on the
positron fraction confirms earlier results
\cite{1969ApJ...158..771F,1975ApJ...199..669B,1987ApJ...312..183M,1994ApJ...436..769G,Barwick:1997ig,Beatty:2004cy,Aguilar:2007yf}
and can be explained by \DM\ annihilation
\cite{2002PhLB..536..263K,2004PhRvD..69j3509H,2008arXiv0810.5344C}. There
is still a good chance that these anomalies can be obviated by
conventional astrophysical effects, e.g., pulsars
\cite{1995A&A...294L..41A,2008arXiv0810.1527H,2008arXiv0810.2784Y,2008arXiv0812.4457P}. Nevertheless,
a plethora of models has been proposed in the last few months
explaining the PAMELA and ATIC results in terms of \DM\ annihilation
into electron-positron pairs (e.g.,
\cite{2008arXiv0810.5557H,ArkaniHamed:2008qn,2008arXiv0811.0399F,2008arXiv0811.1555I,2008arXiv0811.3357C,2008arXiv0811.3641C,2008arXiv0812.2196A}),
a description which also naturally accounts for other experimental
anomalies, such as the ``WMAP haze''
\cite{Finkbeiner:2003im,Dobler:2007wv,Hooper:2007kb}, EGRET
observations of an excess of high-energy gamma-rays in the Galactic
center \cite{Strong:2005zx}, and recent HESS measurements of the flux
of very high-energy electrons \cite{2008arXiv0811.3894H}. Model
building is complicated by the fact that PAMELA did not observe a
similar excess in the ratio of anti-protons to protons
\cite{2008arXiv0810.4994A}, which means that the coupling of the \DM\
particles to quarks must somehow be suppressed.

Of course, the largest obstacle currently to modeling the PAMELA/ATIC
signal with \DM\ annihilation is the very large boost factors ($10^4$
-- $10^5$) with respect to the fiducial $\langle \sigma v
\rangle_{\mathrm{ann}} \approx 3 \times 10^{-26} $ cm$^3$ s$^{-1}$
needed to lead to an observable annihilation flux. The boost factor
from the clumpiness of the dark matter halo is expected to be small
\cite{2008A&A...479..427L,2008Natur.454..735D,2008ApJ...686..262K,2008Natur.456...73S},
typically of the order 1 -- 10. Therefore, a different mechanism must
be responsible for the large boosts necessary. One such mechanism that
has received much attention lately is the Sommerfeld enhancement
\cite{sommerfeld31a,2003PhRvD..67g5014H,2004PhRvL..92c1303H,2005PhRvD..71f3528H,2005PhRvD..71a5007H,2006PhRvD..73e5004H,2008NuPhB.800..204C,2008JHEP...07..058M,2008arXiv0812.0559M,2008arXiv0812.0360L}. The
Sommerfeld enhancement is a generic effect present whenever there is
an attractive force acting between the dark matter particles, \eg, due
to a Yukawa interaction or a gauge interaction through vector bosons
\cite{ArkaniHamed:2008qn}. The main effect of this attractive force is
to enhance the annihilation cross section $\sigma v$ with a factor
proportional to $\beta^{-1}$ ($\beta \equiv v/c$), this is known as
``$1/v$'' enhancement. There are regions in the parameter space made
up of the dark matter mass \mdm, the force carrier mass \mv, and the
force ``fine-structure constant'' $\alpha \equiv$ coupling$^2$/$4 \pi$
in which the enhancement is near a resonance, where it approximately
behaves as $\beta^{-2}$. However, the enhancement levels off at a
certain $\beta$ due to the finite range of the attractive force ---
large ($\alpha \sim\!10^{-1}$ -- $10^{-3}$) attractive forces with
infinite range lead to a burst of \DM\ annihilation in the first \DM\
halos formed at $z$ $\sim$ 100 -- 200 which would lead to observable
effects incompatible with measurements of the diffuse extragalactic
gamma-ray background today and the cosmic microwave background
\cite{2008arXiv0810.3233K}. Since the typical velocities of dark
matter particles in the Galactic halo today are of the order of the
velocity dispersion of the halo, which is non-relativistic ($\beta\sim
10^{-3}$), the cross section of dark matter particles annihilating in
the halo could be significantly enhanced by this Sommerfeld
enhancement.

Both observations
\cite{1994Natur.370..194I,1999Natur.402...53H,2005ApJ...626L..85W,2007ApJ...654..897B,koposov}
as well as numerical simulations
\cite{1998MNRAS.300..146G,johnston98a,1999ApJ...524L..19M,1999ApJ...522...82K,1999MNRAS.307..495H,2003MNRAS.339..834H,2008Natur.454..735D,2008MNRAS.391.1685S}
have firmly established that dark matter halos in \CDM\ are not smooth
structures, but that they are filled with a large amount of
substructure. These substructures are both in the form of subhalos,
self-bound objects orbiting in the Galactic halo, as well as in
streams, relics of tidally disrupted subhalos. Since these
substructures are overdense regions in the halo, the dark matter
annihilation cross section will be correspondingly higher, since
annihilation cross sections scale as the density squared. The effects
of these kinds of substructure boosts have been studied extensively
during the last decade
\cite{2004PhRvD..70b3512B,2004PhRvD..69d3501K,2007ApJ...657..262D,2007PhRvD..75h3526S,2008A&A...479..427L,2008MNRAS.384.1627P,2008ApJ...678..614S,2008ApJ...686..262K,2008Natur.456...73S,2008arXiv0809.2781G}. However,
these subhalos are not only denser than the Galactic halo in which
they orbit, they are also kinematically colder than the Galactic halo,
with velocity dispersions that are typically two orders of magnitude
smaller \cite{1998ARA&A..36..435M,2007ApJ...670..313S}. This means
that the Sommerfeld enhancement in these subhalos can be many orders
of magnitude larger than the Sommerfeld enhancement acting in the
Galactic halo.

In addition to this larger boost due to the small
velocity dispersion in the subhalos, these subhalos themselves are
filled with an abundance of substructure which is approximately
self-similar to that of the main halo. Each of these even smaller
substructure has a larger density than the subhalo they are orbiting
in (quantified by the well established concentration-mass
relationships for \CDM\ halos
\cite{2001MNRAS.321..559B,2001ApJ...554..114E}) as well as a smaller
velocity dispersion. This opens up the possibility that the dark
matter annihilation signal in these already cold subhalos can be
boosted even further by contribution of their substructure.

In this paper we quantify the magnitude of this new substructure boost
and discuss its implications for the detection of dark matter
annihilation from subhalos in the Galactic halos. As we will show, the
total combined boost from Sommerfeld enhancement and substructure can
reach very large values (10$^5$ -- 10$^9$), and these boosts are only
weakly dependent on the assumptions that we have to make about the
abundance of substructure and the \DM\ density profile. We use these
total boosts together with published \DM\ density profiles of some of
the dwarf Spheroidal satellite galaxies of the Milky Way to predict
the total flux of gamma-rays from \DM\ annihilation that would be
detectable by the Fermi Gamma-ray Space Telescope (\Fermi) and find
that in most of the models that posit the Sommerfeld enhancement the
\DM\ annihilation signal from the dwarf Spheroidals shines above the
diffuse extragalactic gamma-ray background, which is the only
background signal at the high Galactic latitudes of these satellite
galaxies. Finally, we also consider the prospects for detecting
smaller subhalos of the Galaxy by computing annihilation fluxes of
subhalos in the \VL\ II simulation. We will see that whether these
subhalos will be detectable in the near future is highly dependent on
the exact details of the Sommerfeld enhancement.

@@

The simplest example of the Sommerfeld effect that contains all of the
generic features described in the introduction is that of the
Sommerfeld enhancement due to an attractive Yukawa type force mediated
by a spin-0 boson $\phi$ between the \DM\ particles, and we adopt that
description here. The Sommerfeld effect is a non-perturbative effect
in the sense that in the language of quantum field theory it comes
about from Feynman diagrams in which the force carrier is exchanged
many times before the annihilation actually happens, giving rise to so
called ``ladder diagrams''. However, we can study the Sommerfeld
effect in simple, non-relativistic quantum mechanics (see the appendix
in Ref.~\cite{ArkaniHamed:2008qn}). We are instructed to solve the
radial Schr{\"o}dinger equation (in natural units)
\begin{equation}\label{eq:sommdiff1}
\frac{1}{\mdm} \frac{\dd^2 \psi(r)}{\dd r^2} + \frac{\alpha}{r}e^{-\mv r} \psi(r) = -\mdm \beta^2 \psi(r)\, ,
\end{equation}
subject to the boundary condition $\dd\psi/\dd r = i \mdm \beta \psi$
as $r \rightarrow \infty$. Here $\mdm$ and $\mv$ are the mass of the
dark matter and the force carrier, respectively. Using the
substitution $r \rightarrow \alpha \mdm r$ this becomes
\begin{equation}\label{eq:sommdiff2}
\frac{\dd^2 \psi(r)}{\dd r^2} + \frac{1}{r}e^{-\mv r/(\alpha \mdm)} \psi(r) = -\frac{\beta^2}{\alpha^2} \psi(r)\, ,
\end{equation}
with boundary condition
\begin{equation}
\psi \propto e^{i\beta r/\alpha} \qquad \mbox{ as } r \rightarrow\infty\, .
\end{equation}
The Sommerfeld enhancement \somm\ is then given by
\begin{equation}
\somm = \frac{|\psi(\infty)|^2}{|\psi(0)|^2}\, .
\end{equation}
From \eqnname\ (\ref{eq:sommdiff2}) it can be seen that the Sommerfeld
enhancement is only a function of two variables, which can be chosen
as $\alpha/\beta$ and $\alpha \mdm/\mv$. The qualitative features of
\eqnname\ (\ref{eq:sommdiff2}) have been widely discussed before and
are easily derived, and, therefore, we will merely mention them here
without proof (see, \eg,
Refs.~\cite{ArkaniHamed:2008qn,2008arXiv0812.0360L}). When the kinetic
energy of the collision is much larger than the product of the force
carrier mass and the coupling, then the Yukawa interaction can be
approximated as an attractive Coulomb interaction, for which an
analytical solution in terms of hypergeometric functions exists. The
Sommerfeld enhancement is simply given by
\begin{equation}
\somm = \frac{\pi \alpha}{\beta} (1 - e^{-\pi\alpha/\beta})^{-1}\, .
\end{equation}
From this we see why the Sommerfeld enhancement is often referred to
as a ``$1/v$'' enhancement for small velocities $\beta \ll 1$. In the
limit of a vanishing interaction ($\alpha/\beta \ll 1$) the Sommerfeld
enhancement disappears as $\somm \rightarrow 1$. The $1/v$ behavior,
which diverges for small $\beta$ and leads to an overproduction of
high-energy photons in the early Universe \cite{2008arXiv0810.3233K},
levels off due to the finite range of the Yukawa interaction at a
velocity $\beta \approx \mv/\mdm$ and settles on a value $\somm \approx
\alpha \mdm/\mv$. However, at some parameter combinations the Yukawa
potential has bound states, at which there are large boosts to the
annihilation cross section which go approximately as $\somm \propto
\beta^{-2}$. These resonance regions, however, are cut-off at small
$\beta$ due to the finite lifetime of the bound state. We expect these
three different behaviors ($1/v$, saturation at low $v$, and resonance
effects due to the existence of bound states) to be generic among
different models of the Sommerfeld enhancement (see also the
discussion in the appendix of Ref.~\cite{ArkaniHamed:2008qn}).

@@

\Eqnname\ (\ref{eq:sommdiff2}) can also be numerically integrated
efficiently, by integrating the solution $\psi(r) = e^{i\beta
r/\alpha}$ inwards from a radius satisfying
\begin{equation}
\frac{1}{r}e^{-\mv r/(\alpha \mdm)} \ll \frac{\beta^2}{\alpha^2} \, .
\end{equation}
In the Coulomb regime ($\mdm \beta^2 \gg \alpha \mv$) this means that
the starting value should be such that $r \gg \alpha^2/\beta^2$;
otherwise we should start at $r \gg \alpha \mdm/\mv$. Figure
\ref{fig:sommb34} shows the Sommerfeld enhancement in the parameter plane
made up off $\alpha$ and $\mv/\mdm$. All of the qualitative features
we discussed above are present: In the Coulomb regime (lower part of
the two panels) we see that the enhancement scales as $\alpha$,
leveling off at $\somm \approx 1$ when $\alpha$ becomes small; The
fact that the upper left corner of both panels is approximately the
same shows that the Sommerfeld enhancement saturates at $\beta \approx
\mv/\mdm$; The sharp diagonal lines in the upper right of both panels
is the resonance region, in which the enhancement is sharply
peaked. Because of the scaling properties of \eqnname\
(\ref{eq:sommdiff2}) shifting $\log\beta$ is equivalent to shifting
both $\log\alpha$ and $\log \mv/\mdm$ by the same amount. Therefore,
the $\beta$ behavior of the enhancement can also be read off from
these figures along diagonals. Inspection of the resonance diagonals
in the second panel of Figure \ref{fig:sommb34} shows that the enhancement
grows as $\beta^{-2}$ in these regions.

Of course, when considering the Sommerfeld enhancement of a halo of a
certain velocity dispersion $\sigv$ we need to average the Sommerfeld
enhancement over the distribution of relative velocities of that
halo. We will approximate the one-particle velocity distribution of
the halo as a single truncated Maxwell-Boltzmann distribution with
velocity dispersion $\sigv$
\begin{equation}\label{eq:maxboltz}
f(v) \propto \left\{
\begin{array}{rl}
v^2\, e^{-v^2/2\sigv^2}  & \qquad v \leq v_{\mathrm{esc}}\\
0 & \qquad v > v_{\mathrm{esc}}\, .\\
\end{array} \right.
\end{equation}
This approximation of a single Maxwell-Boltzmann
distribution for the velocity distribution means that we assume that
the halo has a constant velocity dispersion over the halo, \ie, that
the halo is isothermal. This is not necessarily the case ---indeed, it
is not if we believe that the density profiles of \DM\ halos are well
fit by Einasto or \NFW\ profiles--- but it is not a bad assumption for
intermediate distances from the halo's center, \ie, between the
innermost and outermost parts of the halo
\cite{2007ApJ...667L..53W}. A more detailed treatment of the
Sommerfeld enhancement in subhalos should take the non-uniform
velocity dispersion in subhalos into account. Note that in the pure
$1/v$ regime the averaging will tend to sustain the $1/v$ behavior up
to a factor $\order(1)$, as we have that
\begin{equation}
\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2}
\somm(v)\left/\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2} \right.\propto
\frac{1}{\sigv}\, ,
\end{equation}
since the truncation does not significantly influence the
result. Similarly, in the resonance region the $1/v^2$ behavior is
conserved by the averaging. When the enhancement has leveled off the
averaging has no effect.

>>>>

\begin{abstract}
The recently introduced Sommerfeld enhancement of the dark matter
annihilation cross section has important implications for the
detection of dark matter annihilation in subhalos in the Galactic
halo. In addition to the boost to the dark matter annihilation cross
section from the high densities of these subhalos with respect to the
main halo, an additional boost caused by the Sommerfeld enhancement
results from the fact that they are kinematically colder than the
Galactic halo. If we further believe the generic prediction of \CDM\
that in each subhalo there is an abundance of substructure which is
approximately self-similar to that of the Galactic halo, then I show
that additional boosts coming from the density enhancements of these
small substructures and their small velocity dispersions enhance the
dark matter annihilation cross section even further. I find that very
large boost factors ($10^5$ to $10^9$) are obtained in a large class
of models. The implications of these boost factors for the detection
of dark matter annihilation from dwarf Spheroidal galaxies in the
Galactic halo are such that, generically, they outshine the background
gamma-ray flux and are detectable by the Fermi Gamma-ray Space
Telescope.
\end{abstract}

@@

One exciting possibility is that dark matter self-annihilates at a
level observable in the Galactic halo today. Detecting this
self-annihilation seems unlikely considering the standard
self-annihilation cross section; $\sigmaannv_{\mathrm{ann}} \lesssim 3
\times 10^{-26} $ cm$^3$ s$^{-1}$ for consistency with the observed
relic density. However, there are various mechanisms to boost the
annihilation signal to a value larger by a few orders of magnitude,
e.g., clumpiness of the dark halo
\cite{1993ApJ...411..439S,1999PhRvD..59d3506B,2008A&A...479..427L,2008ApJ...686..262K}
or a combination of coannihilations and Sommerfeld enhancement
\cite{2005PhRvD..72j3521P,2008arXiv0812.0360L}. The possibility of
detecting the \DM\ annihilation signal has attracted much attention
recently because of the anomalous excesses of high energy electrons
and positrons reported by two different experimental groups: the
Payload for Antimatter Exploration and Light-nuclei Astrophysics
(PAMELA) satellite, which saw a sharp upturn in the positron fraction
$e^+/(e^++e^-)$ from 10 to 100 GeV \cite{Adriani:2008zr}, and the
Advanced Thin Ionization Calorimeter (ATIC) group, which reported an
excess over the expected background in the combined number of
electrons and positrons at energies 500 to 800 GeV
\cite{2008Natur.456..362C}. The PAMELA result on the positron fraction
confirms earlier results
\cite{1969ApJ...158..771F,1975ApJ...199..669B,1987ApJ...312..183M,1994ApJ...436..769G,Barwick:1997ig,Beatty:2004cy,Aguilar:2007yf}
and can be explained by \DM\ annihilation
\cite{2002PhLB..536..263K,2004PhRvD..69j3509H,2008arXiv0810.5344C}. There
is still a good chance that these anomalies can be obviated by
conventional astrophysical effects, e.g., pulsars
\cite{1995A&A...294L..41A,2008arXiv0810.1527H,2008arXiv0810.2784Y,2008arXiv0812.4457P}. Nevertheless,
a plethora of models has been proposed in the last few months
explaining the PAMELA and ATIC results in terms of \DM\ annihilation
into electron-positron pairs (e.g.,
\cite{2008arXiv0810.5557H,ArkaniHamed:2008qn,2008arXiv0811.0399F,2008arXiv0811.1555I,2008arXiv0811.3357C,2008arXiv0811.3641C,2008arXiv0812.2196A}),
a description that also naturally accounts for other experimental
anomalies, such as the ``WMAP haze''
\cite{Finkbeiner:2003im,Dobler:2007wv,Hooper:2007kb}, EGRET
observations of an excess of high-energy gamma-rays in the Galactic
center \cite{Strong:2005zx}, and recent HESS measurements of the flux
of very high-energy electrons \cite{2008arXiv0811.3894H}. Model
building is complicated by the fact that PAMELA did not observe a
similar excess in the ratio of anti-protons to protons
\cite{2008arXiv0810.4994A}, which means that the coupling of the \DM\
particles to quarks must somehow be suppressed.

The largest obstacle to modeling the PAMELA/ATIC signal with \DM\
annihilation is the very large boost factor ($10^4$ to $10^5$) with
respect to the fiducial $\langle \sigma v \rangle_{\mathrm{ann}}
\approx 3 \times 10^{-26} $ cm$^3$ s$^{-1}$ required to produce an
observable annihilation flux. The boost factor from the clumpiness of
the dark matter halo is expected to be small
\cite{2008A&A...479..427L,2008Natur.454..735D,2008ApJ...686..262K,2008Natur.456...73S},
typically of the order 1 to 10. Therefore, a different mechanism must
be responsible for the large boosts necessary. One such mechanism that
has received much attention lately is the Sommerfeld enhancement
\cite{sommerfeld31a,2003PhRvD..67g5014H,2004PhRvL..92c1303H,2005PhRvD..71f3528H,2005PhRvD..71a5007H,2006PhRvD..73e5004H,2008NuPhB.800..204C,2008JHEP...07..058M,2008arXiv0812.0559M,2008arXiv0812.0360L}. The
Sommerfeld enhancement is a generic effect present whenever there is
an attractive force acting between the dark matter particles, \eg, a
Yukawa interaction or a gauge interaction through vector bosons
\cite{ArkaniHamed:2008qn}. The main effect of this attractive force is
to enhance the annihilation cross section $\sigma v$ with a factor
proportional to $\beta^{-1}$ ($\beta \equiv v/c$); this is known as
``$1/v$'' enhancement. There are regions in the parameter space made
up of the dark matter mass \mdm, the force carrier mass \mv, and the
force ``fine-structure constant'' $\alpha \equiv$ coupling$^2$/$4 \pi$
in which the enhancement is near a resonance, where it approximately
behaves as $\beta^{-2}$. However, the enhancement saturates at a
certain $\beta$ because of the finite range of the attractive
force---large ($\alpha \sim\!10^{-1}$ to $10^{-3}$) attractive forces
with infinite range lead to a burst of \DM\ annihilation in the first
\DM\ halos formed at $z$ $\sim$ 100 -- 200 which would lead to
observable effects incompatible with measurements of the diffuse
extragalactic gamma-ray background today and the cosmic microwave
background \cite{2008arXiv0810.3233K}. Since the typical velocities of
dark matter particles in the Galactic halo today are of the order of
the velocity dispersion of the halo, which is non-relativistic
($\beta\sim 10^{-3}$), the cross section of dark matter particles
annihilating in the halo could be significantly enhanced by this
Sommerfeld effect.

Both observations
\cite{1994Natur.370..194I,1999Natur.402...53H,2005ApJ...626L..85W,2007ApJ...654..897B,koposov}
as well as numerical simulations
\cite{1998MNRAS.300..146G,johnston98a,1999ApJ...524L..19M,1999ApJ...522...82K,1999MNRAS.307..495H,2003MNRAS.339..834H,2008Natur.454..735D,2008MNRAS.391.1685S}
have firmly established that dark matter halos in \CDM\ are not smooth
structures, but that they are filled with a large amount of
substructure. These substructures are both in the form of subhalos,
self-bound objects orbiting in the Galactic halo, and in the form of
streams, relics of tidally disrupted subhalos. Since these
substructures are overdense regions in the halo, the dark matter
annihilation cross section will be correspondingly higher:
Annihilation cross sections scale as the density squared. The effects
of these kinds of substructure boosts have been studied extensively
during the last decade
\cite{2004PhRvD..70b3512B,2004PhRvD..69d3501K,2007ApJ...657..262D,2007PhRvD..75h3526S,2008A&A...479..427L,2008MNRAS.384.1627P,2008ApJ...678..614S,2008ApJ...686..262K,2008Natur.456...73S,2008arXiv0809.2781G}. However,
these subhalos are not only denser than the Galactic halo in which
they orbit, they are also kinematically colder than the Galactic halo,
with velocity dispersions that are typically two orders of magnitude
smaller \cite{1998ARA&A..36..435M,2007ApJ...670..313S}. This means
that the Sommerfeld enhancement in these subhalos can be many orders
of magnitude larger than the Sommerfeld enhancement acting in the
Galactic halo taken as a whole.

In addition to this larger boost caused by the small velocity
dispersion in the subhalos, these subhalos themselves are filled with
an abundance of substructure which is approximately self-similar to
that of the main halo. Each of these even smaller substructure has a
larger density than the subhalo in which they are orbiting (quantified
by the well established concentration-mass relationships for \CDM\
halos \cite{2001MNRAS.321..559B,2001ApJ...554..114E}) as well as a
smaller velocity dispersion. This opens up the possibility that the
dark matter annihilation signal in these already cold subhalos can be
boosted even further by contribution of their substructure.

In this paper we quantify the magnitude of this new substructure boost
and discuss its implications for the detection of dark matter
annihilation from subhalos in the Galactic halos. As we will show, the
total combined boost from Sommerfeld enhancement and substructure can
reach very large values (10$^5$ to 10$^9$), and these boosts are only
weakly dependent on the assumptions that we have to make about the
abundance of substructure and the \DM\ density profile. We use these
total boosts together with published \DM\ density profiles of some of
the dwarf Spheroidal satellite galaxies of the Milky Way to predict
the total flux of gamma-rays from \DM\ annihilation that would be
detectable by the Fermi Gamma-ray Space Telescope (\Fermi) and find
that in most of the models that posit the Sommerfeld enhancement the
\DM\ annihilation signal from the dwarf Spheroidals shines above the
diffuse extragalactic gamma-ray background, which is the only
background signal at the high Galactic latitudes of these satellite
galaxies. Finally, we also consider the prospects for detecting
smaller subhalos of the Galaxy by computing annihilation fluxes of
subhalos in the \VL\ II simulation. Whether these subhalos will be
detectable in the near future depends on the exact details of the
Sommerfeld enhancement.

@@

The simplest example of the Sommerfeld effect that contains all of the
generic features described in the introduction is that from an
attractive Yukawa type force mediated by a spin-0 boson $\phi$ between
the \DM\ particles, and we adopt that description here. The Sommerfeld
effect is a non-perturbative effect in the sense that in the language
of quantum field theory it comes about from Feynman diagrams in which
the force carrier is exchanged many times before the annihilation
actually happens, giving rise to so called ``ladder
diagrams''. However, we can study the Sommerfeld effect in simple,
non-relativistic quantum mechanics (see the appendix in
Ref.~\cite{ArkaniHamed:2008qn}). We are instructed to solve the radial
Schr{\"o}dinger equation (in natural units)
\begin{equation}\label{eq:sommdiff1}
\frac{1}{\mdm} \frac{\dd^2 \psi(r)}{\dd r^2} + \frac{\alpha}{r}e^{-\mv r} \psi(r) = -\mdm \beta^2 \psi(r)\, ,
\end{equation}
subject to the boundary condition $\dd\psi/\dd r = i \mdm \beta \psi$
as $r \rightarrow \infty$. Here $\mdm$ and $\mv$ are the mass of the
dark matter and the force carrier, respectively. Using the
substitution $r \rightarrow \alpha \mdm r$ this becomes
\begin{equation}\label{eq:sommdiff2}
\frac{\dd^2 \psi(r)}{\dd r^2} + \frac{1}{r}e^{-\mv r/(\alpha \mdm)} \psi(r) = -\frac{\beta^2}{\alpha^2} \psi(r)\, ,
\end{equation}
with boundary condition
\begin{equation}
\psi \propto e^{i\beta r/\alpha} \qquad \mbox{ as } r \rightarrow\infty\, .
\end{equation}
The Sommerfeld enhancement \somm\ is then given by
\begin{equation}
\somm = \frac{|\psi(\infty)|^2}{|\psi(0)|^2}\, .
\end{equation}
\Eqnname~(\ref{eq:sommdiff2}) shows that the Sommerfeld enhancement is
only a function of two variables, which can be chosen as
$\alpha/\beta$ and $\alpha \mdm/\mv$. The qualitative features of
\eqnname\ (\ref{eq:sommdiff2}) have been widely discussed and,
therefore, we will merely mention them here without proof (see, \eg,
Refs.~\cite{ArkaniHamed:2008qn,2008arXiv0812.0360L}). When the kinetic
energy of the collision is much larger than the product of the force
carrier mass and the coupling, then the Yukawa interaction can be
approximated as an attractive Coulomb interaction, for which an
analytical solution in terms of hypergeometric functions exists. The
Sommerfeld enhancement is simply given by
\begin{equation}
\somm = \frac{\pi \alpha}{\beta} (1 - e^{-\pi\alpha/\beta})^{-1}\, .
\end{equation}
From this we see why the Sommerfeld enhancement is often referred to
as a ``$1/v$'' enhancement for small velocities $\beta \ll 1$. In the
limit of a vanishing interaction ($\alpha/\beta \ll 1$) the Sommerfeld
enhancement disappears as $\somm \rightarrow 1$. The $1/v$ behavior,
which diverges for small $\beta$ and leads to an overproduction of
high-energy photons in the early Universe \cite{2008arXiv0810.3233K},
saturates as a consequence of the finite range of the Yukawa
interaction at a velocity $\beta \approx \mv/\mdm$ and settles on a
value $\somm \approx \alpha \mdm/\mv$. However, at some parameter
combinations the Yukawa potential has bound states, at which there are
large boosts to the annihilation cross section which go approximately
as $\somm \propto \beta^{-2}$. These resonance regions, however, are
cut-off at small $\beta$ by the finite lifetime of the bound state. We
expect these three different behaviors ($1/v$, saturation at low $v$,
and resonance effects from the existence of bound states) to be
generic among different models of the Sommerfeld enhancement (see also
the discussion in the appendix of Ref.~\cite{ArkaniHamed:2008qn}).

@@

\Eqnname\ (\ref{eq:sommdiff2}) can also be numerically integrated
efficiently, by integrating the solution $\psi(r) = e^{i\beta
r/\alpha}$ inwards from a radius satisfying
\begin{equation}
\frac{1}{r}e^{-\mv r/(\alpha \mdm)} \ll \frac{\beta^2}{\alpha^2} \, .
\end{equation}
In the Coulomb regime ($\mdm \beta^2 \gg \alpha \mv$) this means that
the initial value for the integration should be such that $r \gg
\alpha^2/\beta^2$; otherwise we should start at $r \gg \alpha
\mdm/\mv$. Figure \ref{fig:sommb34} shows the Sommerfeld enhancement
in the parameter plane made of $\alpha$ and $\mv/\mdm$. All of the
qualitative features we discussed above are present: In the Coulomb
regime (lower part of the two panels) we see that the enhancement
scales as $\alpha$, saturating at $\somm \approx 1$ when $\alpha$
becomes small; The fact that the upper left corner of both panels is
approximately the same shows that the Sommerfeld enhancement saturates
at $\beta \approx \mv/\mdm$; The sharp diagonal lines in the upper
right of both panels show the resonance region, in which the
enhancement is sharply peaked. Because of the scaling properties of
\eqnname\ (\ref{eq:sommdiff2}), shifting $\log\beta$ is equivalent to
shifting both $\log\alpha$ and $\log \mv/\mdm$ by the same
amount. Therefore, the $\beta$ behavior of the enhancement can also be
read off from these figures along diagonals. Inspection of the
resonance diagonals in the second panel of Figure \ref{fig:sommb34}
shows that the enhancement grows as $\beta^{-2}$ in these regions.

Of course, when considering the Sommerfeld enhancement of a halo of a
certain velocity dispersion $\sigv$ we need to average the Sommerfeld
enhancement over the distribution of relative velocities of that
halo. We will approximate the one-particle velocity distribution of
the halo as a single truncated Maxwell-Boltzmann distribution with
velocity dispersion $\sigv$
\begin{equation}\label{eq:maxboltz}
f(v) \propto \left\{
\begin{array}{rl}
v^2\, e^{-v^2/2\sigv^2}  & \qquad v \leq v_{\mathrm{esc}}\\
0 & \qquad v > v_{\mathrm{esc}}\, .\\
\end{array} \right.
\end{equation}
This approximation of a single Maxwell-Boltzmann
distribution for the velocity distribution means that we assume that
the halo has a constant velocity dispersion over the halo, \ie, that
the halo is isothermal. This is not necessarily the case---indeed, it
is not if we believe that the density profiles of \DM\ halos are well
fit by Einasto or \NFW\ profiles---but it is not a bad assumption for
intermediate distances from the halo's center, \ie, between the
innermost and outermost parts of the halo
\cite{2007ApJ...667L..53W}. A more detailed treatment of the
Sommerfeld enhancement in subhalos should take the non-uniform
velocity dispersion in subhalos into account. Note that in the pure
$1/v$ regime the averaging will tend to sustain the $1/v$ behavior up
to a factor $\order(1)$, as we have that
\begin{equation}
\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2}
\somm(v)\left/\int_0^\infty \dd v \, v^2\, e^{-v^2/4\sigv^2} \right.\propto
\frac{1}{\sigv}\, ,
\end{equation}
since the truncation does not significantly influence the result of
these integrals. Similarly, in the resonance region the $1/v^2$
behavior is conserved by the averaging. When the enhancement has
leveled off the averaging has no effect.
